# PBMCs: Data Visualization

## Package Set Up

```{r}
library(dplyr)
library(tidyverse)
library(colorRamp2)
library(ComplexHeatmap)
library(cowplot)
library(ggplot2)
library(ggridges)
library(vegan)
library(scales)
library(colorspace)

```

## Loading in dataframes

```{r}

PT106_test_fisher_categorized <- readRDS("./PT106_test_fisher_categorized.rds")
PT108_test_fisher_categorized <- readRDS("./PT108_test_fisher_categorized.rds")
PT111_test_fisher_categorized <- readRDS("./PT111_test_fisher_categorized.rds")
PT113_test_fisher_categorized <- readRDS("./PT113_test_fisher_categorized.rds")
PT114_test_fisher_categorized <- readRDS( "./PT114_test_fisher_categorized.rds")
PT116_test_fisher_categorized <- readRDS("./PT116_test_fisher_categorized.rds")
PT117_test_fisher_categorized <- readRDS("./PT117_test_fisher_categorized.rds")
PT118_test_fisher_categorized <- readRDS("./PT118_test_fisher_categorized.rds")
PT119_test_fisher_categorized <- readRDS("./PT119_test_fisher_categorized.rds")
PT122_test_fisher_categorized <- readRDS("./PT122_test_fisher_categorized.rds")

```

## Combine Dataframe for Visualization

```{r}
fisher_clusters <- bind_rows(
  PT122_test_fisher_categorized  %>% mutate(patient_id = "PT122"),
  PT119_test_fisher_categorized  %>% mutate(patient_id = "PT119"),
  PT118_test_fisher_categorized  %>% mutate(patient_id = "PT118"),
  PT117_test_fisher_categorized  %>% mutate(patient_id = "PT117"),
  PT116_test_fisher_categorized  %>% mutate(patient_id = "PT116"),
  PT114_test_fisher_categorized  %>% mutate(patient_id = "PT114"),
  PT113_test_fisher_categorized  %>% mutate(patient_id = "PT113"),
  PT111_test_fisher_categorized  %>% mutate(patient_id = "PT111"),
  PT108_test_fisher_categorized  %>% mutate(patient_id = "PT108"),
  PT106_test_fisher_categorized  %>% mutate(patient_id = "PT106")
)

ex <- fisher_clusters %>% 
  group_by(patient_id, new_category) %>%
  summarize(count=n(), .groups = "drop")

```

## Plotting Proportion of Postvax Expanded and Postnivo Expanded

```{r}

library(dplyr)
library(ggplot2)

# Define the relevant categories
sig_categories <- c("post_W14(vaccine)_expanded", "post_W22(booster+Nivo)_expanded")

# Filter only significant clones
sig_clones <- fisher_clusters %>%
  filter(new_category %in% sig_categories)

# Create a new column to label as postvax or postnivo
sig_clones <- sig_clones %>%
  mutate(expansion_type = case_when(
    new_category == "post_W14(vaccine)_expanded" ~ "Postvax",
    new_category == "post_W22(booster+Nivo)_expanded" ~ "Postnivo"
  ))

# Count and calculate proportions per patient
sig_summary <- sig_clones %>%
  group_by(patient_id, expansion_type) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(patient_id) %>%
  mutate(proportion = count / sum(count))

# Set factor levels for patient order
sig_summary$patient_id <- factor(
  sig_summary$patient_id,
  levels = c("PT106","PT108", "PT116", "PT118", "PT119", "PT122", "PT111", "PT113", "PT114", "PT117")
)

# patient groupings
progressive <- c("PT111", "PT114", "PT117", "PT113")
transient_response <- c("PT106", "PT108", "PT116")
stable_disease <- c("PT118", "PT119", "PT122")

# Add response column
sig_summary <- sig_summary %>%
  mutate(response = case_when(
    patient_id %in% progressive ~ "Progressive",
    patient_id %in% transient_response ~ "Transient_Response",
    patient_id %in% stable_disease ~ "Stable_Disease",
    TRUE ~ NA_character_  # For patients not in any group
  ))

# Define color mappings
expansion_colors <- c("Postvax" = "#F8766D", "Postnivo" = "#87CEEB")
response_colors <- c(
  "Progressive" = "#E69F00",
  "Transient_Response" = "#56B4E9",
  "Stable_Disease" = "#009E73"
)


sig_summary$response <- factor(sig_summary$response, levels=c("Transient_Response", "Stable_Disease", "Progressive"))
# Plot with response group as outline color
ggplot(sig_summary, aes(x = patient_id, y = proportion, fill = expansion_type)) +
  geom_bar(stat = "identity", size = 1, position = "stack") +
  labs(
    title = "Proportion of Significant Clones (Postvax vs Postnivo)",
    x = "Patient ID",
    y = "Proportion of Significant Clones",
    fill = "Expansion Type",
    color = "Response Group"
  ) +
  scale_fill_manual(values = expansion_colors) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 1) + facet_grid(~response, scales = "free_x", space = "free_x")


```

## Plotting Number of Postvax Expanded and Postnivo Expanded

```{r, width=24, height = 20}

sig_summary <- sig_summary %>% mutate(new_category = case_when(
    expansion_type == "Postvax" ~ "Post_Vaccine_Expanded",
    expansion_type == "Postnivo" ~ "Post_Nivo_Expanded",
    TRUE ~ expansion_type
  ))

sig_summary$new_category <- factor(sig_summary$new_category, levels=c("Post_Vaccine_Expanded","Post_Nivo_Expanded"))

sig_summary %>%
  ggplot(aes(x = patient_id, y = count, fill = response)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +  # set dodge width
  geom_text(aes(label = count), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5,  # move text slightly above the bar
            size = 4) +
  facet_wrap(~new_category, scales = "free_y") +
  labs(
    title = "Number of Postvaccine and Postnivo Expanded Clones",
    x = "Patient ID",
    y = "Clone Count",
    fill = "Response Group"
  ) +
  scale_fill_manual(
    values = c(
      "Progressive" = "#E69F00",
      "Transient_Response" = "#56B4E9",
      "Stable_Disease" = "#009E73"
    )
  ) +
  theme_classic(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Conducting t-test for the Number of Postvax Expanded vs Postnivo Expanded.

```{r, width=24, height = 20}

# Combine all patient data
all_counts <- bind_rows(
  PT122_test_fisher_categorized  %>% mutate(patient_id = "PT122"),
  PT119_test_fisher_categorized  %>% mutate(patient_id = "PT119"),
  PT118_test_fisher_categorized  %>% mutate(patient_id = "PT118"),
  PT117_test_fisher_categorized  %>% mutate(patient_id = "PT117"),
  PT116_test_fisher_categorized  %>% mutate(patient_id = "PT116"),
  PT114_test_fisher_categorized  %>% mutate(patient_id = "PT114"),
  PT113_test_fisher_categorized  %>% mutate(patient_id = "PT113"),
  PT111_test_fisher_categorized  %>% mutate(patient_id = "PT111"),
  PT108_test_fisher_categorized  %>% mutate(patient_id = "PT108"),
  PT106_test_fisher_categorized  %>% mutate(patient_id = "PT106")
)

# Count clonotypes by category and patient
# Postvax Expanded vs Postnivo Expanded across response groups
clonotype_counts <- all_counts %>%
  group_by(patient_id, new_category) %>%
  summarise(count = n(), .groups = "drop")

# Only interested in postvax and postnivo comparison
paired_data <- clonotype_counts %>%
  filter(new_category %in% c("post_W14(vaccine)_expanded", "post_W22(booster+Nivo)_expanded")) %>%
  pivot_wider(names_from = new_category, values_from = count) %>%
  replace_na(list(
    `post_W14(vaccine)_expanded` = 0,
    `post_W22(booster+Nivo)_expanded` = 0
  ))

# Paired t-test
t_test_result <- t.test(paired_data$`post_W14(vaccine)_expanded`,
                        paired_data$`post_W22(booster+Nivo)_expanded`,
                        paired = TRUE)

# Reshape for plotting
long_counts_df <- paired_data %>%
  pivot_longer(
    cols = c(`post_W14(vaccine)_expanded`, `post_W22(booster+Nivo)_expanded`),
    names_to = "Condition",
    values_to = "Significant_Clones"
  ) %>%
  mutate(
    Condition = recode(Condition,
      "post_W14(vaccine)_expanded" = "Post_W14(Vaccine)",
      "post_W22(booster+Nivo)_expanded" = "Post_W22(Nivo)"
    ),
    Patient = patient_id
  )

# Annotate patient groups
progressive <- c("PT111", "PT114", "PT117", "PT113")
transient_response <- c("PT106", "PT108", "PT116")
stable_disease <- c("PT118", "PT119", "PT122")

long_counts_df <- long_counts_df %>%
  mutate(response = case_when(
    Patient %in% progressive ~ "Progressive",
    Patient %in% transient_response ~ "Transient_Response",
    Patient %in% stable_disease ~ "Stable_Disease",
    TRUE ~ NA_character_
  ))

# Final Plot (faceted by response)
ggplot(long_counts_df, aes(x = Condition, y = Significant_Clones, group = Patient)) +
  geom_point(aes(color = Patient), size = 3, alpha = 0.6) +
  geom_line(aes(color = Patient), size = 1, alpha = 0.6) +
  labs(
    title = "Paired t-test of Significant Clones",
    subtitle = paste("Paired t-test p-value:", signif(t_test_result$p.value, 3)),
    x = "Condition",
    y = "Number of Significant Clones"
  ) +
  facet_wrap(~response, scales = "free_y") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Comparing the Number of Postvax Clone Counts Separated by Response Groups

```{r, width=24, height = 20}

# Filter postvax expanded clones
postvax_counts <- clonotype_counts %>%
  filter(new_category == "post_W14(vaccine)_expanded")

# Label patients by response group
postvax_counts <- postvax_counts %>%
  mutate(response_group = case_when(
    patient_id %in% progressive ~ "Progressive",
    patient_id %in% c(transient_response, stable_disease) ~ "Non_Progressive",
    TRUE ~ NA_character_
  ))

# Perform unpaired t-test
unpaired_t <- t.test(count ~ response_group, data = postvax_counts)

# Plot results
ggplot(postvax_counts, aes(x = response_group, y = count, color = response_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.2) +
  geom_jitter(width = 0.2, size = 3, alpha = 0.6) +
  labs(
    title = "Comparison of Postvax Expanded Clones",
    subtitle = paste("Unpaired t-test p-value:", signif(unpaired_t$p.value, 3)),
    x = "Response Group",
    y = "Number of Postvax Expanded Clones"
  ) +
  scale_color_manual(values = c(
    "Progressive" = "#E69F00",
    "Non_Progressive" = "#56B4E9"
  )) +
  theme_minimal(base_size = 14)


# Filter postvax expanded clones
postvax_counts <- clonotype_counts %>%
  filter(new_category == "post_W22(booster+Nivo)_expanded")

# Label patients by response group
postvax_counts <- postvax_counts %>%
  mutate(response_group = case_when(
    patient_id %in% progressive ~ "Progressive",
    patient_id %in% c(transient_response, stable_disease) ~ "Non_Progressive",
    TRUE ~ NA_character_
  ))

# Perform unpaired t-test
unpaired_t <- t.test(count ~ response_group, data = postvax_counts)

# Plot results
ggplot(postvax_counts, aes(x = response_group, y = count, color = response_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.2) +
  geom_jitter(width = 0.2, size = 3, alpha = 0.6) +
  labs(
    title = "Comparison of Postnivo Expanded Clones",
    subtitle = paste("Unpaired t-test p-value:", signif(unpaired_t$p.value, 3)),
    x = "Response Group",
    y = "Number of Postvax Expanded Clones"
  ) +
  scale_color_manual(values = c(
    "Progressive" = "#E69F00",
    "Non_Progressive" = "#56B4E9"
  )) +
  theme_minimal(base_size = 14)
```

## Heatmap to Show Difference Between Postnivo - Postvaccine Frequencies (among postvaccine expanded clones)

```{r, fig.height=20, fig.width=20}

# color definition function 
col_fun <- colorRamp2(
  breaks = c(-0.01, 0, 0.01),
  colors = c("blue", "white", "red")
)

make_heatmap <- function(df, patient_id) {
  temp <- df[df$new_category == "post_W14(vaccine)_expanded", ]
  temp <- temp[order(-temp$Postvaccine_umi_fraction), ]
  # difference in umi_fraction (Postnivo - Postvaccine)
  temp$diff <- temp$Postnivo_umi_fraction - temp$Postvaccine_umi_fraction
  temp <- temp[order(-temp$diff), ]
  mat <- matrix(temp$diff)
  colnames(mat) <- paste0("Patient:", patient_id)
  Heatmap(mat, cluster_rows = FALSE, cluster_columns = FALSE, show_row_names = FALSE,col=col_fun, heatmap_legend_param = list(title = "Nivo-Vax"))
}

# Create heatmaps for each patient
p1 <- make_heatmap(PT108_test_fisher_categorized, 108)
p2 <- make_heatmap(PT116_test_fisher_categorized, 116)
p3 <- make_heatmap(PT113_test_fisher_categorized, 113)
p4 <- make_heatmap(PT114_test_fisher_categorized, 114)
p5 <- make_heatmap(PT117_test_fisher_categorized, 117)
p6 <- make_heatmap(PT118_test_fisher_categorized, 118)
p7 <- make_heatmap(PT119_test_fisher_categorized, 119)
p8 <- make_heatmap(PT122_test_fisher_categorized, 122)

p1_grid <- grid.grabExpr(draw(p1))
p2_grid <- grid.grabExpr(draw(p2))
p3_grid <- grid.grabExpr(draw(p3))
p4_grid <- grid.grabExpr(draw(p4))
p5_grid <- grid.grabExpr(draw(p5))
p6_grid <- grid.grabExpr(draw(p6))
p7_grid <- grid.grabExpr(draw(p7))
p8_grid <- grid.grabExpr(draw(p8))

plot_grid(p1_grid, p2_grid, p3_grid, p4_grid,p5_grid,p6_grid,p7_grid,p8_grid, ncol=8)

```

## Plotting Frequency of individual clones

```{r}
draw_lod_line <- function(yinter) {
  indiv_lod_line <- geom_hline(yintercept = yinter, linetype = "dashed", color = "red")
  return(indiv_lod_line)
}

plot_frequency <- function(input_df, timepoint_order, p_id) {
  # Long format for plotting
  input_df_pivot <- input_df %>%
    dplyr::select(clonotype_id, contains("umi_fraction")) %>%
    tidyr::pivot_longer(
      cols = dplyr::contains("umi_fraction"),
      names_to = "Timepoint",
      values_to = "UMI_Fraction"
    )
  
  input_df_pivot$Timepoint <- factor(input_df_pivot$Timepoint, levels = timepoint_order)
  input_df_pivot$UMI_Fraction <- input_df_pivot$UMI_Fraction
  input_df_plot_data <- input_df_pivot %>%
    left_join(input_df %>% dplyr::select(clonotype_id, new_category), by = "clonotype_id")
  
  # Count clonotypes per category
  clone_counts <- input_df %>%
    count(new_category) %>%
    deframe()  # turns into a named vector
  
  # Create legend labels with counts
  legend_labels <- c(
    "non_expanded" = paste0("non_expanded (n=", clone_counts["non_expanded"], ")"),
    "post_W14(vaccine)_expanded" = paste0("post_W14(vaccine)_expanded (n=", clone_counts["post_W14(vaccine)_expanded"], ")"),
    "post_W22(booster+Nivo)_expanded" = paste0("post_W22(booster+Nivo)_expanded (n=", clone_counts["post_W22(booster+Nivo)_expanded"], ")")
  )
  
  # Plot
  p1 <- ggplot(input_df_plot_data, aes(x = Timepoint, y = UMI_Fraction, group = clonotype_id, color = new_category)) +
    geom_line(data = input_df_plot_data %>% filter(new_category == "non_expanded"), 
              aes(color = "non_expanded"), size = 1, alpha = 0.5) +
    geom_line(data = input_df_plot_data %>% filter(new_category == "post_W14(vaccine)_expanded"), 
              aes(color = "post_W14(vaccine)_expanded"), size = 1, alpha = 0.5) +
    geom_line(data = input_df_plot_data %>% filter(new_category == "post_W22(booster+Nivo)_expanded"), 
              aes(color = "post_W22(booster+Nivo)_expanded"), size = 1, alpha = 0.5) +
    scale_color_manual(
      values = c(
        "non_expanded" = "grey90",
        "post_W14(vaccine)_expanded" = "darkgoldenrod3",
        "post_W22(booster+Nivo)_expanded" = "#0072B2"
      ),
      labels = legend_labels
    ) +
    labs(title = paste("Clonotype UMI Fraction Across Timepoints", p_id),
         x = "Timepoint",
         y = "UMI Fraction",
         color = "Clone Category") +
    annotation_logticks(sides = "l") +
    draw_lod_line(input_df$visualization_LOD[1]) +
    annotate("text", x = as.numeric(length(timepoint_order)) + 0.25,
             y = input_df$visualization_LOD[1]*2, label = "LOD", size = 4, color = "red") +
    theme_classic(base_size = 15) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      axis.text.y = element_text(size = 8),
      axis.title.x = element_text(size = 8, face = "bold"),
      axis.title.y = element_text(size = 8, face = "bold"),
      legend.text = element_text(size = 8),
      legend.title = element_text(size = 8, face = "bold"),
      plot.title = element_text(size = 10, face = "bold", hjust = 0.5),
      plot.margin = margin(5, 60, 5, 5)
    ) +
    scale_y_log10(labels = label_number(accuracy = 0.00001)) + 
    guides(color = guide_legend(override.aes = list(size = 5, alpha = 1)))

  return(p1)
}
```

## Plotting Line Frequency Plots For Each Patient 

```{r, fig.height=20, fig.width=20}
PT106_timepoint_order <- c("Prevaccine_umi_fraction", "Postvaccine_umi_fraction", "w16_umi_fraction")
PT106_freq <- plot_frequency(PT106_test_fisher_categorized, PT106_timepoint_order, "(PT106)")

PT108_timepoint_order <- c("Prevaccine_umi_fraction", "Postvaccine_umi_fraction", "Postnivo_umi_fraction", "w40_umi_fraction")
PT108_freq <- plot_frequency(PT108_test_fisher_categorized, PT108_timepoint_order, "(PT108)")

PT111_timepoint_order <- c("Prevaccine_umi_fraction", "Postvaccine_umi_fraction")
PT111_freq <- plot_frequency(PT111_test_fisher_categorized, PT111_timepoint_order, "(PT111)")

PT113_timepoint_order <- c("Prevaccine_umi_fraction", "Postvaccine_umi_fraction", "Postnivo_umi_fraction")
PT113_freq <- plot_frequency(PT113_test_fisher_categorized, PT113_timepoint_order, "(PT113)")

PT114_timepoint_order <- c("Prevaccine_umi_fraction", "Postvaccine_umi_fraction", "Postnivo_umi_fraction")
PT114_freq <- plot_frequency(PT114_test_fisher_categorized, PT114_timepoint_order, "(PT114)")

PT116_timepoint_order <- c("Prevaccine_umi_fraction", "Postvaccine_umi_fraction", "Postnivo_umi_fraction", "w40_umi_fraction")
PT116_freq <- plot_frequency(PT116_test_fisher_categorized, PT116_timepoint_order, "(PT116)")

PT117_timepoint_order <- c("Prevaccine_umi_fraction", "Postvaccine_umi_fraction", "Postnivo_umi_fraction")
PT117_freq <- plot_frequency(PT117_test_fisher_categorized, PT117_timepoint_order, "(PT117)")

PT118_timepoint_order <- c("Prevaccine_umi_fraction", "Postvaccine_umi_fraction", "Postnivo_umi_fraction")
PT118_freq <- plot_frequency(PT118_test_fisher_categorized, PT118_timepoint_order, "(PT118)")

PT119_timepoint_order <- c("Prevaccine_umi_fraction", "Postvaccine_umi_fraction", "Postnivo_umi_fraction")
PT119_freq <- plot_frequency(PT119_test_fisher_categorized, PT119_timepoint_order, "(PT119)")

PT122_timepoint_order <- c("Prevaccine_umi_fraction", "Postvaccine_umi_fraction", "Postnivo_umi_fraction")
PT122_freq <- plot_frequency(PT122_test_fisher_categorized, PT122_timepoint_order, "(PT122)")
```

## Plotting Line Plots, Groupping Each Response Groups

```{r, width=20, height=20}
# Create plot grids
#grid1 <- plot_grid(PT111_freq, PT113_freq, PT114_freq, PT117_freq, ncol = 2)
#grid2 <- plot_grid(PT118_freq, PT119_freq, PT122_freq, ncol = 2)
#grid3 <- plot_grid(PT106_freq,PT108_freq, PT116_freq, ncol=3)

PT111_freq
PT113_freq
PT114_freq
PT117_freq
```


```{r, width=40, height=40}
PT118_freq
PT119_freq
PT122_freq
```


```{r, width=40, height=40}
PT106_freq
PT108_freq
PT116_freq
```

## Plotting Frequency Mean of each Clone Category

```{r}


#if there is LATE timepoint, add to the plot as well.
plot_frequency_mean <- function(input_df, timepoint_order,p_id) {
  
  input_df_vis <- input_df
  
  # Add new subcategories for Post_Nivo_Expanded
input_df_vis <- input_df_vis %>%
  mutate(PostVaccine_UMI = Postvaccine_umi,  # Adjust to match your actual column name
         new_category = case_when(
           new_category == "post_W14(vaccine)_expanded" ~ "Post_Vaccine_Expanded",
           new_category == "post_W22(booster+Nivo)_expanded" ~ "Post_Nivo_Expanded"
         ))

# Update factor levels for consistent plotting
input_df_vis$new_category <- factor(input_df_vis$new_category,
                                        levels = c("Post_Vaccine_Expanded",                                           "Post_Nivo_Expanded"))
  input_df_long <- input_df_vis %>%
    dplyr::select(clonotype_id, contains("umi_fraction")) %>%
    tidyr::pivot_longer(
      cols = dplyr::contains("umi_fraction"),
      names_to = "Timepoint",
      values_to = "UMI_Fraction"
    )
  
  input_df_long$Timepoint <- factor(input_df_long$Timepoint, levels = timepoint_order)
  input_df_long$UMI_Fraction <- (input_df_long$UMI_Fraction)
  
  input_df_plot_data <- input_df_long %>%
    left_join(input_df_vis %>% dplyr::select(clonotype_id,response,new_category), by = "clonotype_id")
  
summary_data <- input_df_plot_data %>%
  filter(new_category %in% c("Post_Vaccine_Expanded", "Post_Nivo_Expanded")) %>%
  group_by(new_category, Timepoint) %>%
  summarise(
    mean_UMI = mean(UMI_Fraction, na.rm = TRUE),
    se_UMI = sd(UMI_Fraction, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
)

# Plot with geom_ribbon and geom_line for mean ± SE
p1 <- ggplot(summary_data, aes(x = Timepoint, y = mean_UMI, color = new_category, group = new_category, fill = new_category)) +
  geom_ribbon(aes(ymin = mean_UMI - se_UMI, ymax = mean_UMI + se_UMI), alpha = 0.3, color = NA) +
  geom_line(size = 1) +
  labs(title = paste("Clusters -", p_id),
       x = "Timepoint", y = "Mean UMI Fraction") +
  annotation_logticks(sides = "l") +
  draw_lod_line(input_df_vis$visualization_LOD[1]) +
  annotate("text", x = as.numeric(length(timepoint_order)) + 0.25,
           y = input_df_vis$visualization_LOD[1]*2, 
           label = "LOD", size = 4, color = "red") +
  theme_classic() +
  scale_y_log10(limits = c(0.0001, 0.1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Post_Vaccine_Expanded" = "lightblue", "Post_Nivo_Expanded" = "salmon")) +
  scale_color_manual(values = c("Post_Vaccine_Expanded" = "blue", "Post_Nivo_Expanded" = "red"))
return(p1)
}
```

## Plotting Mean Frequency Plots for each Patient

```{r}
two_timepoint_order <- c("Prevaccine_umis_fraction", "Postvaccine_umi_fraction")
p9 <- plot_frequency_mean( PT106_test_fisher_categorized, two_timepoint_order, "PT106")
p10 <- plot_frequency_mean(PT111_test_fisher_categorized, two_timepoint_order, "PT111")

three_timepoint_order <- c("Prevaccine_umi_fraction", "Postvaccine_umi_fraction", "Postnivo_umi_fraction")
p3 <- plot_frequency_mean( PT122_test_fisher_categorized, three_timepoint_order, "PT122")
p4 <- plot_frequency_mean(PT119_test_fisher_categorized, three_timepoint_order, "PT119")
p5 <- plot_frequency_mean(PT118_test_fisher_categorized, three_timepoint_order, "PT118")


four_timepoint_order <- c("Prevaccine_umi_fraction", "Postvaccine_umi_fraction", "Postnivo_umi_fraction", "w40_umi_fraction")
p1 <- plot_frequency_mean( PT108_test_fisher_categorized, four_timepoint_order, "PT108")
p2 <- plot_frequency_mean(PT116_test_fisher_categorized, four_timepoint_order, "PT116")

p6 <- plot_frequency_mean(PT113_test_fisher_categorized, three_timepoint_order, "PT113")
p7 <- plot_frequency_mean(PT114_test_fisher_categorized, three_timepoint_order, "PT114")
p8 <- plot_frequency_mean(PT117_test_fisher_categorized, three_timepoint_order, "PT117")
```

## Plotting Shannon Diversity for each Patient

```{r}

#Load in unfiltered dataframe -> consider all clones for diversity calculation

PT106_merged_df <- readRDS("./PT106_merged_df.rds")
PT108_merged_df <- readRDS("./PT108_merged_df.rds")
PT111_merged_df <- readRDS("./PT111_merged_df.rds")
PT113_merged_df <- readRDS("./PT113_merged_df.rds")
PT114_merged_df <- readRDS( "./PT114_merged_df.rds")
PT116_merged_df <- readRDS("./PT116_merged_df.rds")
PT117_merged_df <- readRDS("./PT117_merged_df.rds")
PT118_merged_df <- readRDS("./PT118_merged_df.rds")
PT119_merged_df <- readRDS("./PT119_merged_df.rds")
PT122_merged_df <- readRDS("./PT122_merged_df.rds")

calculate_diversity <- function(data_list, patient_name) {
  results <- data.frame()
  
  for (timepoint in names(data_list)) {
    umi_vector <- data_list[[timepoint]]
    if (length(umi_vector) > 1 && sum(umi_vector) > 0) {
      diversity_val <- diversity(umi_vector, index = "shannon") / log10(length(umi_vector))
    } else {
      diversity_val <- NA
    }
    results <- rbind(results, data.frame(
      Patient = patient_name,
      Timepoint = timepoint,
      Normalized_Shannon = diversity_val
    ))
  }
  return(results)
}

# Define timepoint lists for each patient
PT106_list <- list(
  Prevaccine_umi = PT106_merged_df$Prevaccine_umi,
  Postvaccine_umi = PT106_merged_df$Postvaccine_umi,
  w16_umi = PT106_merged_df$w16_umi
)

PT108_list <- list(
  Prevaccine_umi = PT108_merged_df$Prevaccine_umi,
  Postvaccine_umi = PT108_merged_df$Postvaccine_umi,
  Postnivo_umi = PT108_merged_df$Postnivo_umi,
  w40_umi = PT108_merged_df$w40_umi
)

PT111_list <- list(
  Prevaccine_umi = PT111_merged_df$Prevaccine_umi,
  Postvaccine_umi = PT111_merged_df$Postvaccine_umi
)

PT113_list <- list(
  Prevaccine_umi = PT113_merged_df$Prevaccine_umi,
  Postvaccine_umi = PT113_merged_df$Postvaccine_umi,
  Postnivo_umi = PT113_merged_df$Postnivo_umi
)

PT114_list <- list(
  Prevaccine_umi = PT114_merged_df$Prevaccine_umi,
  Postvaccine_umi = PT114_merged_df$Postvaccine_umi,
  Postnivo_umi = PT114_merged_df$Postnivo_umi
)

PT116_list <- list(
  Prevaccine_umi = PT116_merged_df$Prevaccine_umi,
  Postvaccine_umi = PT116_merged_df$Postvaccine_umi,
  Postnivo_umi = PT116_merged_df$Postnivo_umi,
  w40_umi = PT116_merged_df$w40_umi
)

PT117_list <- list(
  Prevaccine_umi = PT117_merged_df$Prevaccine_umi,
  Postvaccine_umi = PT117_merged_df$Postvaccine_umi,
  Postnivo_umi = PT117_merged_df$Postnivo_umi
)

PT118_list <- list(
  Prevaccine_umi = PT118_merged_df$Prevaccine_umi,
  Postvaccine_umi = PT118_merged_df$Postvaccine_umi,
  Postnivo_umi = PT118_merged_df$Postnivo_umi
)

PT119_list <- list(
  Prevaccine_umi = PT119_merged_df$Prevaccine_umi,
  Postvaccine_umi = PT119_merged_df$Postvaccine_umi,
  Postnivo_umi = PT119_merged_df$Postnivo_umi
)

PT122_list <- list(
  Prevaccine_umi = PT122_merged_df$Prevaccine_umi,
  Postvaccine_umi = PT122_merged_df$Postvaccine_umi,
  Postnivo_umi = PT122_merged_df$Postnivo_umi
)

# Apply to all patients
all_diversity <- rbind(
  calculate_diversity(PT106_list, "PT106"),
  calculate_diversity(PT108_list, "PT108"),
  calculate_diversity(PT111_list, "PT111"),
  calculate_diversity(PT113_list, "PT113"),
  calculate_diversity(PT114_list, "PT114"),
  calculate_diversity(PT116_list, "PT116"),
  calculate_diversity(PT117_list, "PT117"),
  calculate_diversity(PT118_list, "PT118"),
  calculate_diversity(PT119_list, "PT119"),
  calculate_diversity(PT122_list, "PT122")
)


```

## Assign Response Groups to Diversity Object and Plot

```{r}
# Base colors for each response group
base_colors <- c(
  "Progressive" = "#E69F00",
  "Transient_Response" = "#56B4E9",
  "Stable_Disease" = "#009E73"
)

# Define patients per group
progressive <- c("PT111", "PT114", "PT117", "PT113")
transient_response <- c("PT106", "PT108", "PT116")
stable_disease <- c("PT118", "PT119", "PT122")

# Function to generate shades for a vector of patients
generate_shades <- function(patients, base_color) {
  n <- length(patients)
  pal <- scales::gradient_n_pal(c(lighten(base_color, 0.3), base_color, darken(base_color, 0.3)))
  shades <- pal(seq(0, 1, length.out = n))
  names(shades) <- patients
  return(shades)
}

# Generate color mapping
patient_colors <- c(
  generate_shades(progressive, base_colors["Progressive"]),
  generate_shades(transient_response, base_colors["Transient_Response"]),
  generate_shades(stable_disease, base_colors["Stable_Disease"])
)

# Example: assign patient colors to the dataframe
all_diversity <- all_diversity %>%
  mutate(patient_color = patient_colors[Patient])

# Add response column
all_diversity <- all_diversity %>%
  mutate(response = case_when(
    Patient %in% progressive ~ "Progressive",
    Patient %in% transient_response ~ "Transient_Response",
    Patient %in% stable_disease ~ "Stable_Disease",
    TRUE ~ NA_character_  # For patients not in any group
  ))

all_diversity$response  <- gsub("_", " ", all_diversity$response)
all_diversity$Timepoint <- gsub("_", " ", all_diversity$Timepoint)
all_diversity$Timepoint <- factor(all_diversity$Timepoint, levels= c(
  "Prevaccine umi",
  "Postvaccine umi",
  "Postnivo umi",
  "w16 umi",
  "w40 umi"
))

p3 <- ggplot(all_diversity, aes(
    x = Timepoint, y = Normalized_Shannon,
    group = Patient, color = Patient
  )) +
  geom_line(alpha = 0.4) +
  geom_point(size = 2, alpha = 0.8) +
  facet_wrap(~response, scales="free_x") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_manual(values = patient_colors) +  # assign colors manually
  labs(y = "Normalized Shannon")

p3
```


```{r}
sessionInfo()
```
